### 2019-03-23 paper/sex_selection_spacing-ver2.tex
Edit Conclusion for grammar

Three major, interrelated trends over the the four decades emerge from the data.
First, fertility has declined for all groups, as shown by the parity progression
probabilities, with a larger decline the more educated the woman.
Across all groups, the likelihood of having an additional child in each spell depends 
strongly on how many sons you already have, with women with none or only one son having
the highest probability of a birth.
Second, there has been a general increase in the length of average birth 
intervals for all education groups.
The increase in average birth interval is larger the higher the spell and the higher the 
education group.
Finally, as shown by the increasing sex ratios for women with no or only one son there
has been a substantial increase in the use of sex selection over time.
As for fertility and birth intervals, this increase is larger the higher the spell and
with higher education.

The results show two very different approaches to son preference with important
impacts on spacing patterns.
At one extreme, rural women without education mostly follow the standard pattern of 
shorter spacing when a woman does not have the desired number of sons.
The strength of this pattern does, however, depend critically on spell number.
For the second spell, although the average interval after a first-born girl is shorter
than the average interval after a first-born boy, the difference is small.
The differences in spacing is larger for the third spell and largest for the fourth.
For the the fourth spell, the divergence in intervals across sex compositions have 
increased over time, with the largest increases in average intervals for women with 
two or three boys.
There has also been an increase in average spacing for women with only girls, and there
is some limited evidence of biased sex ratios, which suggests that this increases in
spacing follow from the use of sex selection.

At the other extreme, urban women with eight or more years of education have experienced 
an almost complete reversal of the traditional spacing patterns as a result of falling
fertility and the associated extensive use of sex selection.
Rather than having the next birth sooner as they mostly did before sex selection became 
available, women with either no or one son now have substantially longer spacing than if 
they have two or more sons.
As shown by the increasingly significant deviations in the sex ratio from the natural sex 
ratio, these changes are likely due to increased use of sex selection rather than sudden
reversal in son preference.
Only the second spell has not seen a reversal despite a substantial increased use of
sex selection, mainly because the average spacing with a first-born boy has increased 
independently by more than nine months.

Despite the differences in average spacing across sex compositions of prior children,
the picture of how son preference affects spacing is more complicated than excepted.
First, for the second spell the differences in birth intervals for the first 25 percent of 
births are negligible across sex compositions, despite that these birth intervals are
clearly within the durations considered the most likely to generate adverse outcomes.
Furthermore, there is little difference in this spacing patterns across education.
Hence, most of the divergence in spacing patterns across sex compositions arise from
longer spells, which are likely to have less negative impacts on children and the mother.
Second, one of the side-effects of the extensive use of sex selection for women with eight
or more years of education for the third and fourth spell is that only-girls families
not only end up with longer average intervals to the birth of the next, they also are now
much less likely to have a short, i.e. less than 18 months, interpregnancy interval,
which may reduce the sibling competition for resources.

With a country as large as India, it is natural to expect variation in both fertility
behavior, spacing patterns, and the degree of son preference.
The regional results confirm this, although differences are less apparent than one would
expect given the prior research on regional differences.
The ``North'' and the ``West'' regions have relatively similar spacing patterns, likely
precipitated by rapidly declining fertility and substantial use of sex selection.
The exception is that the rural areas of the ``North'' still has higher fertility and
consequently less use of sex selection and have, therefore, not yet experienced the
same reversal of spacing patterns.
Even the ``South'', which is otherwise considered the area with the least son preference,
has has seen a substantial increase in the use of sex selection and the associated changes 
in the spacing patterns.
Even though the ``South'' is often mentioned as the low-fertility region of India, 
fertility has declined fastest in the ``East'' region, although this does not appear to 
have translated into the use of sex selection.

The results here lead to a set of broader questions that future research should address.
First, what is behind the substantial increases in spacing even for women who are unlikely 
to use sex selection, and, related, why does spacing appear to increase with declining 
parity progression probabilities?
The distributional results provide an initial indication that most of the increases in 
average birth intervals come from the upper end of the birth intervals rather an generic
increase across the distribution.
It is, however, unclear why we are seeing relatively more very long birth intervals over
time, and the longer average spacing seem to run counter to a general reduction in the 
length of women's reproductive spans found in earlier research \citep{Padmadas2004}.
Understanding whether the longer average spacing is by choice at the beginning of the 
spell, because of unplanned pregnancies, or because of changes in economic circumstances 
that allow families to have an additional child later has important implications for both 
policy design and outcomes for mother and children.

@@

Finally, we need to understand how female labor force participation interacts with the use 
of sex selection.
Increased autonomy for women, arising, for example, from better opportunities for working 
outside the home, has been suggested as a way to increase women's status and thereby lower 
the use of sex selection \citep{Das-Gupta2016}.
This may, however, be a double-edged sword.
On the one hand, it is a clear benefit to the women who gain bargaining power, and it 
increases the cost of repeated sex-selective abortions because the increased duration 
between births would cause a stronger disruption in labor market participation.
On the other hand, it may further lower desired fertility, and that may, everything else 
equal, lead to higher use of sex selection.
Understanding the trade-off between long-term benefits from improvements in women's labor 
force participation and short-term costs from potential increases in sex selection is of 
paramount importance.

>>>>

Three major, interrelated trends over the four decades emerge from the data.
First, fertility has declined for all groups, as shown by the parity progression
probabilities, with a larger decline the more educated the woman.
Across all groups, the likelihood of having an additional child in each spell depends 
strongly on the number of sons, with women with none or only one son having
the highest probability of a birth.
Second, there has been a general increase in the length of average birth 
intervals for all education groups.
The increase in the average birth interval is larger the higher the spell and the higher the 
education group.
Finally, as shown by the increasing sex ratios for women with no or only one son there
has been a substantial increase in the use of sex selection over time.
As for fertility and birth intervals, this increase is larger the higher the spell and
with higher education.

The results show two very different approaches to son preference with significant
impacts on spacing patterns.
At one extreme, rural women without education mostly follow the standard pattern of 
shorter spacing when a woman does not have the desired number of sons.
The strength of this pattern does, however, depend critically on spell number.
For the second spell, although the average interval after a first-born girl is shorter
than the average interval after a first-born boy, the difference is small.
The differences in spacing are larger for the third spell and largest for the fourth.
For the fourth spell, the divergence in intervals across sex compositions has 
increased over time, with the largest increases in average intervals for women with 
two or three boys.
There has also been an increase in average spacing for women with only girls, and there
is some limited evidence of biased sex ratios, which suggests that this increases in
spacing follow from the use of sex selection.

At the other extreme, urban women with eight or more years of education have experienced 
an almost complete reversal of the traditional spacing patterns as a result of falling
fertility and the associated extensive use of sex selection.
Rather than having the next birth sooner as they mostly did before sex selection became 
available, women with either no or one son now have substantially longer spacing than if 
they have two or more sons.
As shown by the increasingly significant deviations in the sex ratio from the natural sex 
ratio, these changes are likely due to increased use of sex selection rather than a sudden
reversal in son preference.
Only the second spell has not seen a reversal despite a substantially increased use of
sex selection, mainly because the average spacing with a first-born boy has increased 
independently by more than nine months.

Despite the differences in average spacing across sex compositions of prior children,
the picture of how son preference affects spacing is more complicated than excepted.
First, for the second spell, the differences in birth intervals for the first 25 percent of 
births are negligible across sex compositions, despite that these birth intervals are
clearly within the durations considered the most likely to generate adverse outcomes.
Furthermore, there is little difference in this spacing patterns across education.
Hence, most of the divergence in spacing patterns across sex compositions arise from
longer spells, which are likely to have less negative impacts on children and the mother.
Second, one of the side-effects of the extensive use of sex selection for third and fourth 
spells is that only-girls families are much less likely to have a short, i.e., less than 
18 months, interpregnancy interval, which may reduce the sibling competition for resources.

With a country as large as India, it is natural to expect variation in both fertility
behavior, spacing patterns, and the degree of son preference.
The regional results confirm this, although differences are less apparent than one would
expect given the prior research on regional differences.
The ``North'' and the ``West'' regions have relatively similar spacing patterns, likely
precipitated by rapidly declining fertility and substantial use of sex selection.
The exception is that the rural areas of the ``North'' still have higher fertility and
consequently less use of sex selection and have, therefore, not yet experienced the
same reversal of spacing patterns.
Even the ``South,'' which is otherwise considered the area with the least son preference,
has seen a substantial increase in the use of sex selection and the associated changes 
in the spacing patterns.
Even though the ``South'' is often mentioned as the low-fertility region of India, 
fertility has declined fastest in the ``East'' region, although this does not appear to 
have translated into the use of sex selection.

The results here lead to a set of broader questions that future research should address.
First, what is behind the substantial increases in spacing even for women who are unlikely 
to use sex selection, and, related, why does spacing appear to increase with declining 
parity progression probabilities?
The distributional results provide an initial indication that most of the increases in 
average birth intervals come from the upper end of the birth intervals rather a general
increase across the distribution.
It is, however, unclear why we are seeing relatively more very long birth intervals over
time, and the longer average spacing seems to run counter to a general reduction in the 
length of women's reproductive spans found in earlier research \citep{Padmadas2004}.
Understanding whether the longer average spacing is by choice at the beginning of the 
spell, because of unplanned pregnancies, or because of changes in economic circumstances 
that allow families to have an additional child later has important implications for both 
policy design and outcomes for mother and children.

@@

Finally, we need to understand how female labor force participation interacts with the use 
of sex selection.
Increased autonomy for women, arising, for example, from better opportunities for working 
outside the home, has been suggested as a way to increase women's status and thereby lower 
the use of sex selection \citep{Das-Gupta2016}.
This approach may, however, be a double-edged sword.
On the one hand, it is a clear benefit to the women who gain bargaining power, and it 
increases the cost of repeated sex-selective abortions because the increased duration 
between births would cause a stronger disruption in labor market participation.
On the other hand, it may further lower desired fertility, and that may, everything else 
equal, lead to higher use of sex selection.
Understanding the trade-off between long-term benefits from improvements in women's labor 
force participation and short-term costs from potential increases in sex selection is of 
paramount importance.
